---
title: "Demo"
author: "Richel J.C. Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(splendid)
```

`splendid` is a SPeciation Likelihood Engine with a DIDapper logo.
(N.B.: `splendid` is a provisional name)

The project is run by G. Laudanno, R.J.C. Bilderbeek and P.M. Santos Neves.

# Goal

Our goal is to make easier to build R packages for likelihood models in macroevolution.

The package is built in a fully modular fashion. In this way the user can build a likelihood package providing:
- one (or more) loglik function(s);
- one (or more) conditioning function(s);
- one (or more) simulation condition(s);
- a function for any event that can occur in simulations;

Once these functions are provided the user should be able to maximize the likelihood and infer the best parameters for any model.

# Use case: birth-death model

First, we'll show how to use `splendid` on a known model: the standard
birth-death model.

This model has two parameters:

 * a birth rate, commonly denoted by $\lambda$
 * an extinction rate, commonly denoted by $\mu$

## Simulating a birth-death phylogeny

In this use case, we will use:

 * speciation rate $\lambda = 1.0$
 * extinction rate $\mu = 0.1$

Or in R:

```{r}
speciation_rate <- 1.0
extinction_rate <- 0.1
```

When simulating a phylogeny, a user needs to specify:

 * when to stop
 * an RNG seed

In this use case, we'll:

 * stop when the phylogeny has twenty taxa
 * restart when the phylogeny has zero taxa
 * an RNG seed of 314

Or in R:

```{r}
#' Determine when a phylogeny simulation must stop
#' @return TRUE if the simulation needs to stop
stop_if_phylogeny <- function(phylogeny) { ape::Ntip(phylogeny) == 20 }

#' Determine when a phylogeny simulation must restart
#' @return TRUE if the simulation needs to be restarted
restart_if_phylogeny <- function(phylogeny) { ape::Ntip(phylogeny) == 0 }

rng_seed <- 314
```
 
How the model works, needs to be specified by a model expert:

 * how a first phylogeny is set up (`sim_initialize_data_new_clade`)
 * how event rates are determined (`sim_sample_deltas`)
 * what happens at such an event (`sim_event`)

In this use case:

 * the first phylogeny is a crown: two taxa that just split 
 * there are two events, thus two rates:
   * specation event rate: $\lamda$ 
   * extinction event rate: simply $\lamda$ 
 * there are two events:
   * a speciaton event: a random node speciates
   * an extintion event: a random node stops

Or in R:

```{r}
#' Create a first starting phylogeny
#' @return a function that creates a phylogeny
create_first_phylogy <- function() { 
 ape::read.tree(text = "(A,B);")
}

#' Calcualate the rates at which all events happen
#' @param phylogeny a phylogeny
#' @param parameters the model's parameters
#' @return a list of events. 
#'   Each element is an event. 
#'   An event has a name (\code{name}) and a rate (\code{rate})
calc_event_rates <- function(phylogeny, parameters) {
  list(
    list(name = "speciation", rate = ape::Ntip(phylogeny) * parameters$lambda),
    list(name = "extinction", rate = ape::Ntip(phylogeny) * parameters$mu)
  )
}

#' Create the phylogeny after an event took place
#' @param phylogeny a phylogeny
#' @param event one event, as created by \link{calc_event_rates}.
#'   An event has a name (\code{name}) and a rate (\code{rate})
#' @return the phylogeny after the event has taken place
process_events <- function(phylogeny, event) {
  if (event$name == "speciation") {
     # Add node to phylogeny
  } else {
     testit::assert(event$name == "extinction")
     # Remove node from phylogeny
  }
  phylogeny 
}
```

Using these functions, `splendid` can simulate a birth-death phylogeny:

```
#' Simulate a macroevolutionary speciation process
#' @param stop_if_phylogeny function that determines 
#'   when a phylogeny simulation must stop. 
#'   The function must have a \code{phylogeny} as an argument,
#'   and return TRUE or FALSE
#' @param restart_if_phylogeny function that determines 
#'   when a phylogeny simulation must restart 
#'   The function must have a \code{phylogeny} as an argument,
#'   and return TRUE or FALSE
#' @param rng_seed a RNG seed
#' @param create_first_phylogy a function to create a first starting
#'   phylogeny. 
#'   The function must have a \code{parameters} as an argument,
#'   and return a phylogeny
#' @param calc_event_rates function to calculate the rates at which events happen
#'   The function must have \code{phylogeny} and \code{parameters} as argument,
#'   and return a list of events.
#'   An event has a name (\code{name}) and a rate (\code{rate})
#' @pparam process_events function to create a phylogeny after an event took place
#'   The function must have \code{phylogeny} and \code{event} as argument,
#'   and return the new phylogeny.
#'   An event has a name (\code{name}) and a rate (\code{rate})
phylogeny <- splendid::sim(
  stop_if_phylogeny = stop_if_phylogeny,
  restart_if_phylogeny = restart_if_phylogeny, 
  create_first_phylogy = create_first_phylogy,
  calc_event_rates = calc_event_rates,
  process_events = process_events,
  rng_seed = rng_seed
)
```

One of the goals of `splendid` is to generalize the simulation a phylogeny.
Although simulating a birth-death process can be done in usually simpler ways,
we are more flexible than ever. And the function can be used to
run newer models, like the SLS model.

## Simulating an SLS phylogeny

To simulate an SLS phylogeny, 

```{r}
#' Create a first starting phylogeny
#' @return a function that creates a phylogeny
create_first_phylogy <- function() { 
 # Something with sim_initialize_data_new_clade
}

#' Calcualate the rates at which all events happen
#' @param phylogeny a phylogeny
#' @param parameters the model's parameters
#' @return a list of events. 
#'   Each element is an event. 
#'   An event has a name (\code{name}) and a rate (\code{rate})
calc_event_rates <- function(phylogeny, parameters) {
  # Something with sim_sample_deltas
}

#' Create the phylogeny after an event took place
#' @param phylogeny a phylogeny
#' @param event one event, as created by \link{calc_event_rates}.
#'   An event has a name (\code{name}) and a rate (\code{rate})
#' @return the phylogeny after the event has taken place
process_events <- function(phylogeny, event) {
  # Something with sim_event
}
```

Using these functions, `splendid` can simulate an SLS phylogeny:

```
#' Simulate a macroevolutionary speciation process
#' @param stop_if_phylogeny function that determines 
#'   when a phylogeny simulation must stop. 
#'   The function must have a \code{phylogeny} as an argument,
#'   and return TRUE or FALSE
#' @param restart_if_phylogeny function that determines 
#'   when a phylogeny simulation must restart 
#'   The function must have a \code{phylogeny} as an argument,
#'   and return TRUE or FALSE
#' @param rng_seed a RNG seed
#' @param create_first_phylogy a function to create a first starting
#'   phylogeny. 
#'   The function must have a \code{parameters} as an argument,
#'   and return a phylogeny
#' @param calc_event_rates function to calculate the rates at which events happen
#'   The function must have \code{phylogeny} and \code{parameters} as argument,
#'   and return a list of events.
#'   An event has a name (\code{name}) and a rate (\code{rate})
#' @pparam process_events function to create a phylogeny after an event took place
#'   The function must have \code{phylogeny} and \code{event} as argument,
#'   and return the new phylogeny.
#'   An event has a name (\code{name}) and a rate (\code{rate})
phylogeny <- splendid::sim(
  stop_if_phylogeny = stop_if_phylogeny,
  restart_if_phylogeny = restart_if_phylogeny, 
  create_first_phylogy = create_first_phylogy,
  calc_event_rates = calc_event_rates,
  process_events = process_events,
  rng_seed = rng_seed
)
```
